\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-03 doc/incertae_sedis_latex2.tex
Update.

\title{\texttt{Supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{
taxa}}
\maketitle

\section{New Introduction}

Supertree methods combine a collection of input trees with different
taxon sets into a single tree on the combined taxon set. While such
methods are usually applied to leaf-labeled phylogenies that have
been estimated from DNA sequence data, we additionally seek to include
taxonomic information. We thus require that one of the input trees
is the taxonomy tree. The taxonomy has a number of features not found
in the input trees, such as having labels for internal nodes, containing
monotypic nodes, and containing \emph{incertae sedis} nodes. We refer
to labels of taxonomy nodes as ``taxa'', and require that labels
and taxonomy nodes have a one-to-one correspondence. We additionally
require that the taxonomy be comprehensive and contain all labels.

One of the problems that arises when attempting to include taxonomic
information is the problem of \emph{incertae sedis }taxa. \emph{Incertae
sedis} taxa are taxa that have unknown taxonomic placement. The common
interpretation of such taxa is that they may not be place further
root-ward, but may be placed within their sibling taxa. This frequently
occurs when a specimen is identified to a taxonomic level such as
the family level but not further. It also occurs when it is unclear
whether a genus forms its own family, or is part of an existing family.
(Mark rewrite?)

@@

The phrase \textquotedblleft \emph{incertae sedis}\textquotedblright ,
which is a Latin phrase meaning ``uncertain seat'', indicates that
a taxon is of uncertain placement. Such groups arise in practice when
they are classified within a parent group, such as Fungi, but the
precise location within the parent group is uncertain. Such a taxon
would then be attached as a child of the parent group, but labeled
as \emph{incertae sedis}.

@@

\paragraph{Background}

@@

@@

We extend this framework by adding a set $\mathcal{U}\subseteq\mathcal{N}$
of labels that have the \emph{incertae sedis }property. The extended
taxonomy then consists of the taxonomy tree $\taxonomy$, the labels
$\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis }taxa.

\subsection{A splits-based semantics for \emph{incertae sedis} taxa}

@@

\subsubsection{Reduced exclude sets allow \emph{incertae sedis} to be placed into
sibling taxa}

An \emph{incertae sedis} taxon can be moved into any of the descendants
of its siblings. Therefore marking a taxon as \emph{incertae sedis}
changes the meaning of its sibling taxa. We seek to represent this
by constructing modified splits for each branch of the taxonomy tree.
The include sets of these splits remain unchanged, but we construct
reduced exclude sets in order to allow \emph{incertae sedis} taxa
to intrude into their sibling taxa.

@@

@@

After a synthesis tree is constructed, we attach taxon names for to
the synthesis tree for each non-conflicting taxon. These names attach
at the MRCA of the include group. Then, for each taxon $A$ on the
synth tree, we associate with the first taxon $B$ found in a rootward
walk on the synthesis tree. If $A$ is not a descendant of $B$ on
the taxonomy, then we say that $A$ is \emph{placed} within $B$ by
the synthesis tree. While this may occur if $A$ is \emph{incertae
sedis}, more complex scenarios are possible. We note that if method
does not handle \emph{incertae sedis} taxa, then this scenario could
not occur because any taxa $B$ that $A$ is placed within would be
considered incompatible with the synthesis tree and so their names
would not be applied.

In order to fully describe a placement of $A$ within $B$ we consider
the sequence of taxa encountered on a walk from $A$ or $B$ up to
their MRCA $M$ on the taxonomy tree. Let us denote the sequence of
taxa encountered from $A$ as $A_{1},\ldots,A_{n},M$ where $A_{1}=A$.
Let $B_{1},\ldots,B_{m},M$ be the sequence of taxa encountered from
$B$ where $B_{1}=B$. Then $A_{n}$ must be \emph{incertae sedis},
and $A_{2},\ldots,A_{n}$ must be incompatible with the synthesis
tree (i.e. broken taxa). We then have that $A$ is placed successively
within $B_{m},B_{m-1},\ldots,B_{2},B_{1}$, and $m$ is the depth
of the placement of $A$.

One question is whether this approach is able to handle nested incertae
sedis taxa.

@@

Each input tree can relate to an \emph{incertae sedis} taxon $(A,B,C)D$
in a number of ways
\begin{itemize}
\item it could resolve $A$, $B$, $C$, or $D$.
\item it could place $D$ on a degree-2 (=out-degree-1) node that bisects
a branch
\item it could place a descendant taxon of $D$ 
\item it could place a descendant taxon of $D$ in a \emph{different place}
than another input tree.
\item it could place children of $D$ in multiple places, thus conflicting
with the branch. If the \emph{incertae sedis} taxon $(A,B,C)$ is
broken, then $A$, $B$ and $C$ become \emph{incertae sedis} clades
in their own right, that may attach separately, except that they .
This is because none of $A$, $B$, or $C$ is in the exclude set
of the siblings of $D$.
\end{itemize}

@@

>>>>

\title{\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae
sedis}}\texttt{ taxa}}
\maketitle

@@

Supertree methods combine a collection of input trees with different
taxon sets into a single tree on the combined taxon set. While such
methods are usually applied to leaf-labeled phylogenies that have
been estimated from DNA sequence data, we seek to include taxonomic
information also. Taxonomies can provide a comprehensive coverage
of known taxa that inferred phylogenies cannot. Additionally, taxonomies
can provide taxonomic names, which are needed to align input phylogenies
to each other, and to allow references to existing groups.

We thus require that one of the input trees is a taxonomy tree. The
taxonomy has a number of features not found in the input phylogenies.
While the input trees need not cover the entire taxon set, we require
the taxonomy to be comprehensive. The taxonomy also has labels for
internal nodes, whereas input phylogenies only have labels for leaf
nodes. We refer to labels of taxonomy nodes as ``taxon names'',
and require that taxon names and taxonomy nodes have a one-to-one
correspondence. The taxonomy may also contain out-degree 1 nodes.
These nodes correspond to monotypic taxa which contain a single child
of lower taxonomic rank.

In practice, taxonomies may also contain nodes with uncertain placement.
These nodes are often labelled ``\emph{incertae sedis}'', which
means ``uncertain seat'' in Latin. The common interpretation of
such taxa is that they may not be moved further towards the root,
but may be moved into their sibling taxa. For example, a genus with
a sibling that is a family may be annotated as``\emph{incertae familia}'',
indicating that it is unclear which family the genus should be placed
in. \emph{Incertae sedis} taxa frequently occur when a specimen is
identified up to a given taxonomic level, but no further. Extinct
taxa are often \emph{incertae sedis}.

@@

\subsection{Background}

\paragraph{Terminology}

@@

In this manuscript we extend this framework by adding a set $\mathcal{U}\subseteq\mathcal{N}$
of labels that have the \emph{incertae sedis }property. The extended
taxonomy then consists of the taxonomy tree $\taxonomy$, the labels
$\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis }taxa.

@@

 We seek a semantics for supertrees with \emph{incertae sedis }taxa
that satisfy the following properties:
\begin{enumerate}
\item An incertae sedis node may intrude into its siblings and their descendants.
\item Two incertae sedis siblings may not be interdigitated.
\item Any placement $T\to T^{*}$ of a tree $T$ with incertae sedis taxa
is also accessible from a tree $T^{\prime}$ that is created from
$T$ by taking a taxon, attaching it to an ancestor, and marking it
\emph{incertae sedis}.
\item The semantics of an \emph{incertae sedis} node does not depend on
assigned ranks, but only on the tree.
\item Semantics is based on deriving a split for each branch of the tree.
\end{enumerate}
Here we focus on a split-based semantics instead of a semantics based
on apply of a series of edit operations to the tree. The split-based
semantics allows us to seek trees that are simultaneously consistent
with a set of split-constraints. The use of a split-based semantics
does impose some limitations. These come primarily because in the
edit-based semantics, after we place an \emph{incertae sedis} taxon,
it is no loner \emph{incertae sedis}, whereas with the split-based
semantics taxa are either always \emph{incertae sedis} or never \emph{incertae
sedis}.

This has an effect, for example, when we consider two incertae sedis
siblings $A$ and $B$. If we can place $A$ within $B$, and $B$
within $A$, then a split-based semantics allows us to freely inter-digitate
$A$ and $B$. If we do not want to allow $A$ and $B$ to be inter-digitated,
then we cannot both $A$ within $B$ and $B$ within $A$. In addition,
we can imagine that a family $A$ is incertae sedis within a kingdom.
In an edit-based semantics, we could place $A$ as a sibling to a
genus $B$, mark $A$ as non-\emph{incertae-sedis}, and then place
$B$ within $A$. In a split-based semantics, we must disallow either
$B$ within $A$ or $A$ within $B$ if we want to disallow interdigitation.

As a result of this, we cannot satisfy all five of these criteria.
Thus, we describe two alternative split-based semantics. The non-interdigitation
semantics gets \# (1,2,4,5), and the interdigitation semantics gets
(1,3,4,5). This makes the point that each semantics results from different
choices about which criteria to satisfy.

\subsection{Split-based semantics for \emph{incertae sedis} taxa}

@@

\subsubsection{A semantics that disallows inter-digitating \emph{incertae sedis}
siblings}

An \emph{incertae sedis} taxon can be moved into any of the descendants
of its siblings. We seek to represent this by constructing modified
splits for each branch of the taxonomy tree. The include sets of these
splits remain unchanged, but we construct reduced exclude sets in
order to allow \emph{incertae sedis} taxa to intrude into their sibling
taxa.

@@

\subsubsection{An alternative semantics that allows inter-digitating incertae sedis
siblings}

If we want to allow placing two incertae sedis siblings within each
other, then incertae sedis siblings are never excluded.

Furthermore, consider placing $A$ as a sibling of $B$, and then
placing $B$ within $A$. In order for this to work, $A$ must not
exclude the incertae sedis children of $B$.

In order to accomplish both of these goals, we define the descendants
$\mathcal{I}^{\prime}(n)$ that are accessible without passing through
an incertae sedis node.

\begin{align*}
\mathcal{I^{\prime}}(n) & =\begin{cases}
\emptyset & \text{if }n\in\mathcal{U}\\
\{n\} & \text{if }n\text{ is a leaf}\\{}
[\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
\end{cases}
\end{align*}
We may then define the taxa excluded from $n$ as:

\begin{align*}
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup[\mathcal{I}^{\prime}(s)|s\in siblings(n)].
\end{align*}

This semantics satisfies criteria 1,3,4, and 5 above.

\subsubsection{A third semantics}

OK, the idea here is that we want to avoid inter-digitation. \emph{Incertae
sedis} taxa thus exclude\emph{ incertae sedis} siblings. However,
where the first semantics allows \emph{incertae sedis }taxa to intrude
into\emph{ incertae sedis }descendants of siblings, but not allow\emph{
incertae sedis} descendants of siblings to intrude into \emph{incertae
sedis} taxa, here we do the reverse.

OK, so the idea here is that an\emph{ incertae sedis} group excludes
\begin{itemize}
\item \emph{incertae sedis} siblings
\item \emph{incertae sedis }descendants of ancestors
\item but not\emph{ incertae sedis} descendants of non-incertae-sedis siblings
\end{itemize}
It is also possible to allow placing $A$ as a sibling of $B$ and
then placing $B$ within $A$ if we disallow placing $A$ into incertae
sedis descendants of its siblings. We may then allow placing those
descendants within $A$.

\begin{align*}
\mathcal{I^{\prime}}(n) & =\begin{cases}
\emptyset & \text{if }n\in\mathcal{U}\\
\{n\} & \text{if }n\text{ is a leaf}\\{}
[\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
\end{cases}
\end{align*}

We use $\mathcal{I}(m,n)$ to indicate the descendants of $m$ that
are excluded from $n$.
\begin{align*}
\mathcal{I}(m,n) & =\begin{cases}
\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{ is not}\\
\mathcal{I}(m) & \text{if }n\text{ and }m\text{ are both }incertae\,sedis\\
\mathcal{I}^{\prime}(m) & \text{otherwise.}
\end{cases}
\end{align*}

We may then define the taxa excluded from $n$ as:

\begin{align*}
\mathcal{E}(n) & =\begin{cases}
\left[\mathcal{L}-\mathcal{I}(parent(n))\right]\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{if }n\in\mathcal{U}\\
\mathcal{E}(parent(n))\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{otherwise.}
\end{cases}
\end{align*}
I think this works... but need to double-check.

Also: can we genericize $\mathcal{E}(parent(n))$?

@@

After a synthesis tree is constructed, we attach taxon names to the
synthesis tree for each non-conflicting taxon. These names attach
at the MRCA of the include group. Then, for each taxon $A$ that is
found on the synth tree, we associate with the first taxon $B$ found
in a rootward walk on the synthesis tree. If $A$ is not a descendant
of $B$ on the taxonomy, then we say that $A$ is \emph{placed} within
$B$ by the synthesis tree. While this may occur if $A$ is \emph{incertae
sedis}, more complex scenarios are possible. We note that if method
does not handle \emph{incertae sedis} taxa, then this scenario could
not occur because any taxa $B$ that $A$ is placed within would be
considered incompatible with the synthesis tree and so their names
would not be applied.

In order to fully describe a placement of $A$ within $B$ we consider
the sequence of taxa encountered on a walk from $A$ or $B$ up to
their MRCA $M$ on the taxonomy tree. Let us denote the sequence of
taxa encountered from $A$ as $A_{1},\ldots,A_{n},M$ where $A_{1}=A$.
Let $B_{1},\ldots,B_{m},M$ be the sequence of taxa encountered from
$B$ where $B_{1}=B$. Then $A_{n}$ must be \emph{incertae sedis},
and $A_{2},\ldots,A_{n}$ must be incompatible with the synthesis
tree (i.e. broken taxa). We then have that $A$ is placed successively
within $B_{m},B_{m-1},\ldots,B_{2},B_{1}$, and $m$ is the depth
of the placement of $A$.%
\begin{comment}
One question is whether this approach is able to handle nested incertae
sedis taxa.

@@

Each input tree can relate to an \emph{incertae sedis} taxon $(A,B,C)D$
in a number of ways
\begin{itemize}
\item it could resolve $A$, $B$, $C$, or $D$.
\item it could place $D$ on a degree-2 (=out-degree-1) node that bisects
a branch
\item it could place a descendant taxon of $D$ 
\item it could place a descendant taxon of $D$ in a \emph{different place}
than another input tree.
\item it could place children of $D$ in multiple places, thus conflicting
with the branch. If the \emph{incertae sedis} taxon $(A,B,C)$ is
broken, then $A$, $B$ and $C$ become \emph{incertae sedis} clades
in their own right, that may attach separately, except that they .
This is because none of $A$, $B$, or $C$ is in the exclude set
of the siblings of $D$.
\end{itemize}
\end{comment}

@@

With a splits-based semantics, we cannot allow both (i) placing $A$
in $B$ and (ii) placing $B$ in $A$ and also forbid interdigitating
$A$ and $B$.
